\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-12-22 nobuild/jss_revision/response.tex

\begin{response}
The descriptions of the two sparsity patterns are correct. The
first index in the subscript of $\beta$ refers to a household, and the
second index refers to a covariate. Thus,
$\parD{2}{\beta_{11},\beta_{12}}{}\neq 0$, but
$\parD{2}{\beta_{11},\beta_{21}}{}=0$.  When the second index changes
faster than the first, the pattern is ``block-arrow.''  When the first
index changes faster than the second, the pattern is ``banded.''

\end{response}

\item \begin{revQuote}
Code in the paper prior to table 4 needs to be copied from the
vinettes/sparseHessianFD.Rnw file. For those trying to reproduce results it
would be nice if this where mentioned in the file replication.R.
  \end{revQuote}

  \begin{response}
Assuming that the R package builder works the way I think it does,
the \filename{replication.R} file, and the data for Table 4 (\filename{vignette\_tab4.Rdata}), will be
copied to the \filename{doc\textbackslash} directory of the package
once it is installed.  It will also be available as an online supplement to
this paper.
\begin{itemize}
\item 
\end{itemize}

Meeting the documentation and replication needs for both the journal,
and CRAN, can be tricky, but I think I have a good solution

  Note to me:  I think what this means is that results for Table 4 and
  Figure 4 are "hard coded.''  For the jss article, the replication
  file is provided, but this file is not part of the package.  It
  should be, so non-jss vignette readers know how to replicate the results. 
\end{response}

@@

\begin{response}

Here I am combining my responses to items \ref{it:complex1} and
\ref{it:complex2} together.

  The complex step function is now implemented in the package, and is
  introduced in Section 2.5 of the paper.  You
  are correct that ``most R primitives work with complex numbers,''
  but there are many commonly used functions that do not, including 
  \func{log1p}, \func{gamma}, and the probability distribution
  functions. This limitation, combined with the requirement that the function be
  holomorphic, and my own inexperience with the complex step method, make
  me hesitant to promote it too heavily in the paper. But it is a nifty
  trick, and I am happy that it is now included in the package.

In terms of accuracy, the paper now includes comparisons between the true
Hessian and the two \class{sparseHessianFD} methods (finite differences
and complex step). To maintain the focus of the paper on
\pkg{sparseHessianFD}, I am not including an assessment of the accuracy of
\pkg{numDeriv}.  However, for the
timing comparisons in Table 4, I did
change the baseline \pkg{numDeriv} method from \func{hessian} with the
\method{Richardson} method, to \func{jacobian} with both the
\method{simple} and 
\method{complex} methods. Removing the second-order approximation saves
a lot of time, and lets me compute Table 4 without resorting to
parallel computation. That will simplify attempts at replication, and
removes another variable that might influence results.

@@

really serves to highlight the improvement of the sparse calculation because
it is a valid comparison. Even though the results are not as exaggerated,
they are still important:

(My laptop is a  Intel(R) Core(TM) i5-3337U CPU @ 1.80GHz,  4GB RAM, SSD
swap, running Mint variant of Ubuntu 14.04.2 LTS.)

(Code and output removed)
  \end{revQuote}

  \begin{response}
    I agree.  See my response to Item \ref{it:complex2}.  Table 4
    includes times from the \pkg{numDeriv}
    \method{complex} and \method{simple} methods.

    To reduce clutter in Table 4, I removed comparisons of the
    hessian-to-gradient computation time ratio.  That was not a
    particularly interesting metric.

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.tex
almost done with revision

To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in
milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables.  The setup time for
\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
were collected using the \pkg{microbenchmark} package \citep{R_microbenchmark}.

\begin{table}[ptb]\centering
  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }cccp{2pt}rrrrp{2pt}rrrr}
  \toprule
&&&&\multicolumn{4}{c}{\bfseries{Hessian
    time}}&&\multicolumn{4}{c}{\bfseries{Hessian/Gradient ratio}}\\
&&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
$N$&$k$&$M$&&mean&sd&mean&sd&&mean&sd&mean&sd\\
  \cmidrule{1-3}\cmidrule{5-8}\cmidrule{10-13}
   9 &    2 &   20 &&  363 & 6.4 &  2.8 & 2.2 &&  871 & 63 & 6.5 & 4.2 \\
    12 &    2 &   26 &&  629 &  8.9 &  2.9 & 2.1 && 1469 &  124 & 6.5 & 3.3 \\
     9 &    3 &   30 &&  819 & 27.7 &  3.6 & 2.2 && 1974 &  161 & 8.4 & 3.7 \\
    15 &    2 &   32 &&  973 & 25.8 &    3.0 & 2.2 && 2191 &  209 &  6.5 & 3.4 \\
    12 &    3 &   39 && 1407 & 39.3 & 3.5 & 2.0 && 3291 &  285 &    8.0 & 3.5 \\
     9 &    4 &   40 && 1445 & 28.8 & 4.3 & 2.1 && 3453 &  292 & 9.9 & 3.1 \\
    15 &    3 &   48 && 2138 & 92.7 & 3.5 & 2.0 && 4861 &  341 & 7.8 & 3.4 \\
    12 &    4 &   52 && 2400 & 98.6 & 4.3 & 2.1 && 5679 &  503 & 10.1 & 3.7 \\
    15 &    4 &   64 && 3643 &  209.0 & 4.2 & 2.2 && 8945 &  828 & 10.0 &  2.8 \\
   \bottomrule
\end{tabularx}
\caption{Computation times (milliseconds) for computing Hessians using the
  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, across 200 replications. Rows are ordered by the
  number of variables.}\label{tab:numDeriv}
\end{table}

@@

\begin{table}[ptb]\centering
  \begin{tabularx}{1.0\linewidth}{>{\bfseries}l X}
    \toprule
    Measure&Description\\
    \midrule
Function&estimating the objective function\\
Gradient&estimating the gradient\\
Hessian&computing the Hessian (not including initialization or
  partitioning time)\\
Partitioning&finding a consistent partitioning of the
  variables (the vertex coloring problem)\\
Initialization&total setup time (including the partitioning time)\\
Hessian/Gradient&ratio of the Hessian time to the
    gradient time\\
    \bottomrule
  \end{tabularx}
  \caption{Summary of timing tests (see Figure~\ref{fig:timing}).}\label{tab:timeMeasures}
\end{table}

In the plots in Figure~\ref{fig:timing}, the number of heterogeneous units ($N$) is
on the x-axis, and mean run time, in milliseconds, is on the y-axis.  Each panel shows the
relationship between $N$ and run time for a different measure of time,
and each curve in a panel represents a different number of within-unit
parameters ($k$).

@@

Computation times
for the function and gradient, as well as the setup and
partitioning times for the \class{sparseHessianFD} object, grow
linearly with the number of heterogenous units.  The time
for the Hessian grows linearly as well, and that might be partially
surprising.  We saw in Section \ref{sec:coloring} that adding
additional heterogeneous units in a hierarchical model does not
increase the number of required gradient evaluations.  So we might
think that the time
to compute a Hessian should not increase with $N$ \emph{at all}.  The
reason it does is that each gradient evaluation takes longer.  The
plot of the ratio of Hessian-to-gradient run times is relatively flat as the size
of the dataset increases.  We can conclude that the
\pkg{sparseHessianFD} algorithms are quite efficient and scalable for
hierarchical models.

>>>>

To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in
milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables. Times
were collected using the \pkg{microbenchmark} package
\citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
is in the file \filename{replication.R}, which is available as an
online supplement to this paper.

\begin{Schunk}
\begin{Sinput}
R> library(xtable)
R> library(dplyr)
R> load("vignette_tab4.Rdata")
\end{Sinput}
\end{Schunk}

\begin{table}[ptb]\centering
  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
  \toprule
&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
$N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
%%\midrule
  \midrule
15 & 2 & 32 & 12.4 & 1.2 & 2.3 & 0.5 & 12.8 & 1.2 & 2.1 & 0.5 \\ 
  15 & 4 & 64 & 26.2 & 12.4 & 4.0 & 0.9 & 26.1 & 2.0 & 3.8 & 1.1 \\ 
  50 & 2 & 102 & 50.5 & 2.4 & 3.0 & 0.9 & 54.8 & 3.5 & 2.9 & 2.1 \\ 
  15 & 8 & 128 & 50.4 & 2.2 & 8.5 & 12.0 & 55.0 & 2.7 & 7.7 & 1.5 \\ 
  100 & 2 & 202 & 134.1 & 15.7 & 3.7 & 0.5 & 144.0 & 11.6 & 3.3 & 0.6 \\ 
  50 & 4 & 204 & 104.0 & 2.1 & 6.3 & 11.6 & 117.2 & 17.6 & 5.1 & 1.0 \\ 
  100 & 4 & 404 & 284.3 & 34.3 & 8.0 & 10.9 & 322.5 & 40.8 & 8.2 & 11.1 \\ 
  50 & 8 & 408 & 248.5 & 48.7 & 10.9 & 1.4 & 271.7 & 43.7 & 11.5 & 1.4 \\ 
  100 & 8 & 808 & 624.5 & 54.0 & 14.4 & 1.8 & 762.6 & 76.9 & 15.7 & 1.7 \\ 
  500 & 2 & 1002 & 1758.6 & 86.7 & 10.1 & 10.0 & 1990.0 & 90.6 & 9.3 & 9.8 \\ 
  500 & 4 & 2004 & 3832.5 & 84.9 & 18.4 & 2.3 & 4460.7 & 82.8 & 19.4 & 2.5 \\ 
  500 & 8 & 4008 & 8846.5 & 82.1 & 45.2 & 6.2 & 11058.4 & 101.3 & 52.5 & 6.3 \\ 
   \bottomrule

\end{tabularx}
\caption{Computation times (milliseconds) for computing Hessians using the
  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, and the finite
 difference and complex step methods, across 200 replications. Rows are ordered by the
  number of variables.}\label{tab:numDeriv}
\end{table}

@@

\begin{table}[ptb]\centering
  \begin{tabularx}{1.0\linewidth}{>{\bfseries}l X}
    \toprule
    Measure&Description\\
    \midrule
Function&estimating the objective function\\
Gradient&estimating the gradient\\
Hessian&computing the Hessian (not including initialization or
  partitioning time)\\
Partitioning&finding a consistent partitioning of the
  variables (the vertex coloring problem)\\
Initialization&total setup time (including the partitioning time)\\
    \bottomrule
  \end{tabularx}
  \caption{Summary of timing tests (see Figure~\ref{fig:timing}).}\label{tab:timeMeasures}
\end{table}

In the plots in Figure~\ref{fig:timing}, the number of heterogeneous units ($N$) is
on the x-axis, and median run time, in milliseconds, is on the y-axis.  Each panel shows the
relationship between $N$ and run time for a different measure of time,
and each curve in a panel represents a different number of within-unit
parameters ($k$).

@@

Computation times
for the function and gradient, as well as the setup and
partitioning times for the \class{sparseHessianFD} object, grow
linearly with the number of heterogenous units.  The time
for the Hessian grows linearly as well, and that might be partially
surprising.  We saw in Section \ref{sec:coloring} that adding
additional heterogeneous units in a hierarchical model does not
increase the number of required gradient evaluations.  So we might
think that the time
to compute a Hessian should not increase with $N$ \emph{at all}.  The
reason it does is that each gradient evaluation takes longer.
Nevertheless, we can conclude that the
\pkg{sparseHessianFD} algorithms are quite efficient and scalable for
hierarchical models.
